Predicting complex diseases: performance, robustness and ..utility..(impact?)

### 2018-02-27 5-results-celiac.Rmd
need to finish real data

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", fig.width = 10,
                      fig.asp = 0.7, dev = "svg")
options(width = 95)
```

@@

```{r}
# GWAS
gwas.train <- big_univLogReg(
  G, y[ind.train], ind.train = ind.train, 
  covar.train = obj.svd$u[ind.train, , drop = FALSE], 
  ncores = NCORES
)
snp_qq(gwas.train) +
  coord_cartesian(ylim = c(0, 25))
gwas.train.gc <- snp_gc(gwas.train)
```

```{r, dev="png"}
labels <- 1:22; labels[c(11, 13, 15, 17, 18, 20, 21)] <- ""
cowplot::plot_grid(
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels),
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels) +
    coord_cartesian(ylim = c(0, 25)) + 
    geom_hline(yintercept = -log10(5e-8), color = "red", linetype = 3),
  align = "hv", ncol = 1, labels = LETTERS[1:2], label_size = 25, scale = 0.95
)
```

@@

```{r}
# PRS
thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
lpS <- -predict(gwas.train.gc)
prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
               ind.test = ind.test,
               ind.keep = ind.keep,
               lpS.keep = lpS[ind.keep], 
               thr.list = thrs)
nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
aucs <- apply(prs, 2, AUC, target = y[ind.test])
```

@@

```{r}
logit <- big_spLogReg(X = G, y01.train = y[ind.train], 
                      ind.train = ind.train, 
                      covar.train = obj.svd$u[ind.train, ],
                      alpha = 0.5, dfmax = 20e3)
preds1 <- predict(logit, X = G, ind.row = ind.test, 
                  covar.row = obj.svd$u[ind.test, ])
library(Matrix)
nb.pred1 <- colSums(logit$beta != 0)
aucs1 <- apply(preds1, 2, AUC, target = y[ind.test])
```

@@

```{r}
system.time({
  cmsa.logit <- big_CMSA(FUN = big_spLogReg, feval = AUC,
                         X = G, y.train = y[ind.train], 
                         ind.train = ind.train, 
                         covar.train = obj.svd$u[ind.train, ],
                         alpha = 0.5, dfmax = 20e3, 
                         ncores = NCORES)

  preds2 <- predict(cmsa.logit, X = G, ind.row = ind.test, 
                    covar.row = obj.svd$u[ind.test, ])
})
(auc.cmsa <- AUC(preds2, y[ind.test]))
```

@@

```{r}
G2 <- big_attach("backingfiles/celiacQC_tripled1.rds")
logit2 <- big_spLogReg(X = G2, y01.train = y[ind.train], 
                       ind.train = ind.train, 
                       covar.train = obj.svd$u[ind.train, ],
                       alpha = 0.5, dfmax = 20e3)
preds3 <- predict(logit2, X = G2, ind.row = ind.test, 
                  covar.row = obj.svd$u[ind.test, ])
nb.pred2 <- colSums(logit2$beta != 0)
aucs2 <- apply(preds3, 2, AUC, target = y[ind.test])
```

```{r}
system.time({
  cmsa.logit2 <- big_CMSA(FUN = big_spLogReg, feval = AUC,
                          X = G2, y.train = y[ind.train], 
                          ind.train = ind.train, 
                          covar.train = obj.svd$u[ind.train, ],
                          alpha = 0.5, dfmax = 20e3, 
                          ncores = NCORES)

  preds4 <- predict(cmsa.logit2, X = G2, ind.row = ind.test, 
                    covar.row = obj.svd$u[ind.test, ])
})
(auc.cmsa2 <- AUC(preds4, y[ind.test]))
```

@@

First, we get the SNP with the highest ratio of cases over controls for homozygous-recessive individuals.

```{r}
props <- sapply(cols_along(G), function(j) {
  ccoj <- counts.control[, j]
  mut <- `if`(ccoj[1] < ccoj[3], 1, 3)
  `if`(counts.case[mut, j] > 100, counts.case[mut, j] / ccoj[mut], 0)
})
theone <- which.max(props)
table(Geno = G[, theone], forcats::fct_recode(as.factor(y), Control = "0", Case = "1"))
```

```{r}
# one geno
ind0 <- which(G[ind.test, theone] == 0)
(M <- length(ind0))
mean(y[ind.test[ind0]]) 
# prs max
indPRS <- order(prs[, which.max(aucs)], decreasing = TRUE)[1:M]
mean(y[ind.test[indPRS]])
# logit-simple (CMSA)
indLS <- order(preds2, decreasing = TRUE)[1:M]
mean(y[ind.test[indLS]])
# logit-triple (CMSA)
indLT <- order(preds4, decreasing = TRUE)[1:M]
mean(y[ind.test[indLT]])
```

@@

  ind.train <- sort(sample(nrow(G)))

@@

  saveRDS(cmsa.logit[cols_along(G)], file = res.file)
}
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", fig.width = 10,
                      fig.asp = 0.7, dev = "svg", cache = TRUE)
options(width = 95)
```

@@

```{r}
# GWAS
system.time(
  gwas.train <- big_univLogReg(
    G, y[ind.train], ind.train = ind.train, 
    covar.train = obj.svd$u[ind.train, , drop = FALSE], 
    ncores = NCORES
  )
)
```

```{r}
# Q-Q plot
snp_qq(gwas.train) +
  coord_cartesian(ylim = c(0, 20))
gwas.train.gc <- snp_gc(gwas.train)
```

```{r, dev="png"}
# Manhattan plot
labels <- 1:22; labels[c(11, 13, 15, 17, 18, 20, 21)] <- ""
cowplot::plot_grid(
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels),
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels) +
    coord_cartesian(ylim = c(0, 22)) + 
    geom_hline(yintercept = -log10(5e-8), color = "red", linetype = 3),
  align = "hv", ncol = 1, labels = LETTERS[1:2], label_size = 25, scale = 0.95
)
```

@@

```{r}
# PRS
thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
lpS <- -predict(gwas.train.gc)
prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
               ind.test = ind.test,
               ind.keep = ind.keep,
               lpS.keep = lpS[ind.keep], 
               thr.list = thrs)
nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
aucs <- apply(prs, 2, AUC, target = y[ind.test])
summary(aucs)
```

```{r}
pAUC <- function(pred, target, p = 0.1) {
  val.min <- min(target)
  q <- quantile(pred[target == val.min], probs = 1 - p)
  ind <- (target != val.min) | (pred > q)
  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
           error = function(e) p^2 / 2)  # all same prediction
}
plot_density_AUC <- function(pred, true = y[ind.test]) {
  data.frame(pred = pred, pheno = as.factor(true)) %>%
  myggplot() +
  geom_density(aes(pred, fill = pheno), alpha = 0.4) +
  ggtitle(sprintf("AUC = %s%% | pAUC = %s", 
                  round(100 * AUC(pred, true), 2),
                  round(pAUC(pred, true), 4)))
}
plot_density_AUC(prs[, which.max(aucs)])
```

@@

@@

```{r}
# Train logit-simple
system.time(
  logit <- big_spLogReg(X = G, y01.train = y[ind.train], 
                        ind.train = ind.train, 
                        covar.train = obj.svd$u[ind.train, ],
                        ncores = NCORES)
)
# Get K = 10 predictions/individuals
preds1 <- predict(logit, X = G, ind.row = ind.test, 
                  covar.row = obj.svd$u[ind.test, ])
(nb.pred1 <- sum(rowSums(sapply(logit, function(l) l$beta.X != 0)) > 0))
(aucs1 <- apply(preds1, 2, AUC, target = y[ind.test]))
# Combining all K predictions
preds1.comb <- rowMeans(preds1)
AUC(preds1.comb, y[ind.test])
```

```{r}
plot_density_AUC(preds1.comb)
```

@@

```{r}
G2 <- big_attach("backingfiles/celiacQC_tripled1.rds")
system.time(
  logit2 <- big_spLogReg(X = G2, y01.train = y[ind.train], 
                         ind.train = ind.train, 
                         covar.train = obj.svd$u[ind.train, ],
                         ncores = NCORES)
)

preds2 <- predict(logit2, X = G2, ind.row = ind.test, 
                  covar.row = obj.svd$u[ind.test, ])
(nb.pred2 <- sum(rowSums(sapply(logit2, function(l) l$beta.X != 0)) > 0))
(aucs2 <- apply(preds2, 2, AUC, target = y[ind.test]))
# Combining all K predictions
preds2.comb <- rowMeans(preds2)
AUC(preds2.comb, y[ind.test])
```

```{r}
plot_density_AUC(preds2.comb)
```

@@

First, we get the most significant SNP.

```{r}
theone <- which.max(abs(gwas.train.gc$score))
ind0 <- which(G[ind.test, theone] == 0)
(M <- length(ind0))
table(Geno = G[, theone], forcats::fct_recode(as.factor(y), Control = "0", Case = "1"))
```

```{r}
# one geno
mean(y[ind.test[ind0]]) 
# prs max
indPRS <- order(prs[, which.max(aucs)], decreasing = TRUE)[1:M]
mean(y[ind.test[indPRS]])
# logit-simple (CMSA)
indLS <- order(preds2, decreasing = TRUE)[1:M]
mean(y[ind.test[indLS]])
# logit-triple (CMSA)
indLT <- order(preds4, decreasing = TRUE)[1:M]
mean(y[ind.test[indLT]])
```

@@

  ind.train <- sort(sample(nrow(G), replace = TRUE))

@@

  saveRDS(list(betas = cmsa.logit[cols_along(G)],
               ind.test = setdiff(rows_along(G), ind.train)),
          file = res.file)
}
```

### 2018-02-27 paper/draft.tex
need to finish real data

\newpage
\begin{table}[htbp]
\caption{Summary of all simulations. When there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
\vspace*{0.5em}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
scenario & & training set & (number and location) & of effects & & model & \\
\hline
\hline
\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{simple} & PRS \\
& & & 30 in all & & & & logit-simple \\
& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{fancy} & logit-triple \\
& & & 3000 in all & & & & (T-Trees) \\
\hline
\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{simple} & PRS \\ 
& & & & & & & logit-simple\\
\hline
\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
& & 2000 & & & & & \\
& & 3000 & & & & & \\
& & 4000 & & & & & \\
& & 5000 & & & & & \\
\hline
\end{tabular}
\end{table}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-celiac-roc}}
\caption{ROC Curves for the C+T, ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.}
\label{fig:supp-celiac-roc}
\end{figure}

>>>>

\newpage
\begin{table}[htbp]
\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
\vspace*{0.5em}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
scenario & & training set & (number and location) & of effects & & model & \\
\hline
\hline
\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{simple} & PRS \\
& & & 30 in all & & & & logit-simple \\
& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{fancy} & logit-triple \\
& & & 3000 in all & & & & (T-Trees) \\
\hline
\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{simple} & PRS \\ 
& & & & & & & logit-simple\\
\hline
\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
& & 2000 & & & & & \\
& & 3000 & & & & & \\
& & 4000 & & & & & \\
& & 5000 & & & & & \\
\hline
\end{tabular}
\end{table}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-celiac-roc}}
\caption{ROC Curves for the ``C+T'', ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.}
\label{fig:supp-celiac-roc}
\end{figure}
